Covering Your Basis: Model Agnostic Probes of the Binary Black Hole Population

### 2022-09-26 src/tex/results.tex
move figs
add tilt spin results text -- chieff only left

%PLPeak: $m_{1\%} = $\result{$\CIPlusMinus{\macros[MassDistribution][PLPeak][m_1percentile]}$}, $m_{99\%} = $\result{$\CIPlusMinus{\macros[MassDistribution][PLPeak][m_99percentile]}$}
%PLSpline: $m_{1\%} = $\result{$\CIPlusMinus{\macros[MassDistribution][PLSpline][m_1percentile]}$}, $m_{99\%} = $\result{$\CIPlusMinus{\macros[MassDistribution][PLSpline][m_99percentile]}$}
%MSpline: $m_{1\%} = $\result{$\CIPlusMinus{\macros[MassDistribution][MSpline][m_1percentile]}$}, $m_{99\%} = $\result{$\CIPlusMinus{\macros[MassDistribution][MSpline][m_99percentile]}$}

We apply our newly constructed models to the BBH analysis that was done in the astrophysical population companion paper to GWTC-3. 
Figure \ref{fig:mass_distribution} shows how the MSpline model (red) found consistent features in the mass disribution with 
what was inferred with the \textsc{PowerlawPeak} and \textsc{PowerlawSpline} mass models \cite{Talbot_2019,o3a_pop,Edelman_2022ApJ,o3b_astro_dist}. 
In particular our MSpline model finds a peak in merger rate density at both 10 and 35 soloar masses, agreeing with those reported using the same dataset in \citet{o3b_astro_dist}. 
We also find a significantly less declining mass distribution that has considerably larger support for higher mass mergers when 
compared to the other power-law based models. Another interesting consequence of this is that we no longer find an confidently
evolving merger rate with redshift, as inferred with power-law based models. There is a strong correlation between the primary mass 
power law index, $\alpha$, and the redshift power-law index, $\lambda_z$. When one infers a steeper mass distribution -- the inferred 
rate must drastically increase with resdshift to explain the highest mass events in the catalog. However, when one infers a 
flatter distribution like, as the MSPline did in figure \ref{fig:mass_distribution}, the higher mass mergers are consistent with 
the local merger rate denstiy as a function of mass. With the more flexible mass population model, we find that the merger rate 
is consistent with being unifom with co-moving volume with $\lambda_z = $\result{$\CIPlusMinus{\macros[MSplineIIDCompSpins][lamb]}$}, compared 
to $\lambda_z = $\result{$\CIPlusMinus{\macros[PLPeak][lamb]}$} found with the \textsc{PowerlawPeak} model.

We find a similar mass ratio distribution to what was found with the found with the \textsc{PowerlawPeak} model in \citet{o3b_astro_dist}. 
With the MSpline model on primary mass the mass ratio power law index is $\beta = $\result{$\CIPlusMinus{\macros[MSplineIIDCompSpins][beta]}$}, 
compared to $\beta = $\result{$\CIPlusMinus{\macros[PLPeak][beta]}$} for \textsc{PowerlawPeak}.

@@

@@

@@

MSpline IID:
$\cos{\theta}_\mathrm{max} = $\result{$\CIPlusMinus{\macros[MSplineIIDCompSpins][peakCosTilt]}$}
$\gamma = $\result{$\CIPlusMinus{\macros[MSplineIIDCompSpins][log10gammaFrac]}$}
$f_{<0} = $\result{$\CIPlusMinus{\macros[MSplineIIDCompSpins][negFrac]}$}
MSpline Independent:
$\cos{\theta_1}_\mathrm{max} = $\result{$\CIPlusMinus{\macros[MSplineIndependentCompSpins][peakCosTilt1]}$} and $\cos{\theta_2}_\mathrm{max} = $\result{$\CIPlusMinus{\macros[MSplineIndependentCompSpins][peakCosTilt2]}$}
$\gamma_1 = $\result{$\CIPlusMinus{\macros[MSplineIndependentCompSpins][log10gammaFrac1]}$} and $\gamma_2 = $\result{$\CIPlusMinus{\macros[MSplineIndependentCompSpins][log10gammaFrac2]}$} 
$f_{1,<0} = $\result{$\CIPlusMinus{\macros[MSplineIndependentCompSpins][negFrac1]}$} andand $f_{2,<0} = $\result{$\CIPlusMinus{\macros[MSplineIndependentCompSpins][negFrac2]}$}

The spins distribution of merging combact binaries provides a very useful insight into the formation history of the binary. In particular the two main subcategories, 
the isolated and dynamical formation channels, would produce distinct signatures in the spin distribution when looking at a population of observations. Isolated formation
is where the only interactions with the binary are between each other. As the binary evolves and mass transfer happens between the two components, we expect the orientation
of each components spin vector to be preferentially aligned with the orbital angular mommentumof the system. Some misalignment is epected to occur due to the kicks from 
each star exploding in supernovae leaving behind the compact objects. Conversely, the dynamical channel     
\begin{figure}
    \script{plot_iid_spinmag.py}
    \includegraphics[width=0.45\textwidth]{figures/iid_spinmag.pdf}
    \caption{Caption.}
    \label{fig:iid_spinmag_dist}
\end{figure}

@@

Comment here are the inferences of iid model vs Default and then iid vs independent test pointing out possible reasons for why spin tilt 
does not peak at cos theta = 1 i.e. salvo's paper

>>>>

We apply our newly constructed models to the BBH analysis that was done in the astrophysical population companion paper to GWTC-3. 
Figure \ref{fig:mass_distribution} shows how the MSpline model (red) found consistent features in the mass disribution with 
what was inferred with the \textsc{PowerlawPeak} and \textsc{PowerlawSpline} mass models \cite{Talbot_2019,o3a_pop,Edelman_2022ApJ,o3b_astro_dist}. 
In particular our MSpline model finds a peak in merger rate density at both 10 and 35 soloar masses, agreeing with those reported using the same dataset in \citet{o3b_astro_dist}. 
We also find a significantly less declining mass distribution that has considerably larger support for higher mass mergers when 
compared to the other power-law based models, finding that 75\% percent of primary masses below \result{$\CIPlusMinus{\macros[MassDistribution][MSpline][m_75percentile]}$}, 
at 90\% credibility, compared to \result{$\CIPlusMinus{\macros[MassDistribution][PLSpline][m_75percentile]}$}, found with the similar \textsc{PowerlawSpline} model. 
An interesting consequence of this is that we do not find a confidently evolving merger rate with redshift, as inferred with power-law based mass models in \citet{o3b_astro_dist}. 
There is a strong correlation between the primary mass power law index, $\alpha$, and the redshift power-law index, $\lambda_z$. When one infers a steeper mass distribution -- 
the inferred rate must drastically increase with resdshift to explain the highest mass events in the catalog. However, when one infers a 
flatter distribution like, as the MSPline did in figure \ref{fig:mass_distribution}, the higher mass mergers are consistent with 
the local merger rate denstiy as a function of mass given the selection effects. With our more flexible mass model, we find that the merger rate 
is consistent with being unifom with co-moving volume, with $\lambda_z = $\result{$\CIPlusMinus{\macros[MSplineIIDCompSpins][lamb]}$}, compared with 
\citet{o3b_astro_dist}'s reported value -- $\lambda_z = $\result{$\CIPlusMinus{\macros[PLPeak][lamb]}$} with the \textsc{PowerlawPeak} mass model.

We find a similar mass ratio distribution to what was found with the found with the \textsc{PowerlawPeak} model in \citet{o3b_astro_dist}. 
With the MSpline model on primary mass the mass ratio power law index is $\beta = $\result{$\CIPlusMinus{\macros[MSplineIIDCompSpins][beta]}$}, 
compared to $\beta = $\result{$\CIPlusMinus{\macros[PLPeak][beta]}$} for \textsc{PowerlawPeak}. We find that BBH mergers favor more equal mass mergers
as would be expected of both main formation channels due to binary mass transfer in isolated formation and three-body interactions that preferentially eject
the lightest of the three compact objects, in dynamical formation \bruce{cite these}. We also find no evidence for a sharp fall off in merger rate following 
the pileup at $\sim35\,\msun$, which is expected if such a pileup was due to pulsational pair instability supernovae (PPISNe). The lack of any high mass truncation 
along with the peak at 35\msun, significantly lower than expected, may pose challenges for conventional stellar evolution theory and could be a sign of 
the significant presence of sub-populations that may avoid PISN as the binary forms.

@@

\begin{figure}
    \script{plot_iid_spinmag.py}
    \includegraphics[width=0.45\textwidth]{figures/iid_spinmag.pdf}
    \caption{Caption.}
    \label{fig:iid_spinmag_dist}
\end{figure}

@@

The \textsc{Default} spin model models the spin magnitude of both components as identical and indendendently distributed (IID) non-singular Beta distributions. The
Beta distribution provides a simple 2-parameter model that can produce a wide range of functional forms on the unit interval. However, the constraint that keeps 
the Beta distribution non-singluar (i.e. $\alpha>1$ and $\beta>1$) enforces a spin magnitude that always has $p(a_i=0) = 0$. While in reality, it is expected that 
all physical bodies must have at least some non-zero angular mommentum, studies have proposed the existence of a distinct sub-population of non-spinning or 
negliglby spinning black holes that can elude discovery with such a model.

We model the spin magnitude distibutions as IID MSpline distibutions as a flexible check
to the assumptions of the \textsc{Default} model. Figure \ref{fig:iid_spinmag_dist} shows the inferred spin magnitude distribution with the \textsc{MSpline} model, 
compared with the \textsc{Default} model from \citet{o3b_astro_dist}. The MSpline model finds broad agreement with the beta distribution, both peaking near ~0.2, with 90\% 
of BBH spins below \result{$\CIPlusMinus{\macros[MSplineIIDCompSpins][a_90percentile]}$} at 90\% credibility. The regions of parameter space where they don't agree as well are the bounds. 
The MSpline model has no enforcment for vanishing support at the extremal values like the Beta distribution, 
allowing it to probe the zero-spin question. We find broad support, with large variance, for non-zero probabilities at $a_i=0$, but cannot confidently determine the presence of 
a significant non-spinning subpopulation, corroborating the conclusions of \citet{BuildBetterSpinModels,Callister_NoEvidence,GWTC3MonashSpin}. 
We now repeat the same analysis as above but with independent MSpline distibutions for each spin magnitude component. In figure \ref{fig:ind_spinmag_dist} 
we show the inferred primary (orange), and secondary (olive) spin magnitude distributions inferred when relaxing the IID assumption. We find both 
distributions to be remarkably similar, showing no signs that each component spin magnitude are distributed differently from each other.

@@

The \textsc{Default} spin model models the spin orientation of both components as identical and indendendently distributed (IID) mixture models over an
aligned and an isotropic component. The aligned component is modeled with a truncated Gaussian distribution with mean at $\cos{\theta}=1$ and variance a free 
hyper-parameter to be fit. This provides a simple 2-parameter model motivated by simple distributions expected from the two main formation scenario families, allowing 
for a straightforward interpretation of results. One possible limitation however, is that by construction this distribution is forced to peak at perfecty aligned spins, 
i.e. $\cos{\theta}=1$. While this may be a reasonable assumption recently \citet{spinitasyoulike} recently extended the model space of parameteric descriptions 
used to model the spin orientation distribution and found considerably evidence that the distibution peaks away from $\cos{\theta}=1$. Again, this provides a clear 
use-case where data-driven models can help in our understanding and possibly resolve disputes among differing parametric approaches.

@@

We repeat the tests done above on the the spin magnitude distibutions now with the spin orientation distributions. Figure \ref{fig:iid_spintilt_dist} 
shows the inferred spin magnitude distribution with the \textsc{MSpline} model, compared with the \textsc{Default} model from \citet{o3b_astro_dist}. 
The MSpline inferences have large uncertainties but start to show the same features as found and discussed in \citet{spinitasyoulike}. 
We find a distribution that instead of intrinsicly peaking at $\cos{\theta}=1$, is found to peak at: $\cos{\theta}=$\result{$\CIPlusMinus{\macros[MSplineIIDCompSpins][peakCosTilt]}$}, at 
90\% credibility. We find less, but still considerable, support for misaligned spins (i.e. $\cos{\theta}<0$) as the \textsc{Default} model. Specifically we 
find that the fraction of misaligned systems is $f_{<0}=$\result{$\CIPlusMinus{\macros[MSplineIIDCompSpins][negFrac]}$}. This implies 
the presence of an isotropic component as expected by dynamical formation channels, albeit less than with the \textsc{Default} model. To quantify the 
amount of isotropy in the tilt distibution we calculate Y, the ratio defined in Eq. 4 of \citet{spinitasyoulike} which is 1 when the distribution is 
purely isotropic, less than one when negative (anti-aligned) values are favored, and greater than one when positive (aligned) values are favored.
We find a $\log_{10}Y=$\result{$\CIPlusMinus{\macros[MSplineIIDCompSpins][log10gammaFrac]}$}.

Now we model each of the components' orientation distributions with independent \textsc{MSpline} models as done above, and show the inferred 
distributions in figure \ref{fig:ind_spintilt_dist}. The primary and secondary orientation distibutions do not agree as well as the magnitude 
distibutions did in figure \ref{fig:ind_spinmag_dist}, and the PPDs look to peak at different values, but the wide uncertainties still are 
consistent with both each other and the \textsc{Default} model's PPD. We find the two distributions to peak at: $\cos{\theta_1}=$\result{$\CIPlusMinus{\macros[MSplineIndependentCompSpins][peakCosTilt1]}$} 
and $\cos{\theta_2}=$\result{$\CIPlusMinus{\macros[MSplineIndependentCompSpins][peakCosTilt2]}$}, showing that the primary distribution peak is inferred further away 
from the assumed $\cos{\theta}=1$ with the \textsc{Default} model. There is also significant (albeit uncertain) evidence of spin mis-alignments in each distribution, finding 
the fraction of mis-aligned primary and secondary components as: $f_{1,<0}=$\result{$\CIPlusMinus{\macros[MSplineIndependentCompSpins][negFrac1]}$} andand $f_{2,<0}=$\result{$\CIPlusMinus{\macros[MSplineIndependentCompSpins][negFrac2]}$}.
We again calculate the isotropy ratio, Y, for each and find: $\log_{10}Y_1=$\result{$\CIPlusMinus{\macros[MSplineIndependentCompSpins][log10gammaFrac1]}$} and 
$\log_{10}Y_2=$\result{$\CIPlusMinus{\macros[MSplineIndependentCompSpins][log10gammaFrac2]}$}.
